---
title: "Enron Email Analysis"
author: "Nihal Sachindra"
date: "18-05-2025"
output: html_document
---

## Introduction:

This project explores internal email patterns within Enron using R and RStudio.
The study focuses on examining sender behavior, exploring the roles of employees within the organization, analyzing communication trends over time, and investigating the content of email subjects.Through this analysis, I aim to exhibit meaningful insights into organizational dynamics and information flow within Enron.

## Libraries Used:

```{r setup, include=FALSE}

library(dplyr)
library(lubridate)
library(ggplot2)
library(tidytext)

```

## Step 1: Load the Data in RStudio

```{r}
load("Enron (1).Rdata")
ls()
```

## Step 2: Explore the Data

```{r}
str(employeelist)
head(employeelist)
summary(employeelist)
str(message)
head(message)
summary(message)
str(recipientinfo)
head(recipientinfo)
summary(recipientinfo)
glimpse(employeelist)
glimpse(message)
glimpse(recipientinfo)
```


## Step 3: Top Email Senders

```{r step3, echo=FALSE, message=FALSE}

top_senders <- message %>%
  group_by(sender) %>%
  summarise(n_sent = n()) %>%
  arrange(desc(n_sent))
head(top_senders, 10)

top_senders %>%
  slice_max(n_sent, n = 10) %>%
  ggplot(aes(x = reorder(sender, n_sent), y = n_sent)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Most Active Email Senders",
    x = "Sender",
    y = "Number of Emails Sent"
  ) +
  theme_minimal()

```
Insight: A small group of employees generated most of the internal emails.




## Step 4: Email Activity by Role

```{r step4, echo=FALSE, message=FALSE}
message_with_status <- message %>%
  left_join(employeelist, by = c("sender" = "Email_id"))
emails_by_status_clean <- message_with_status %>%
  filter(!is.na(status)) %>%
  group_by(status) %>%
  summarize(n_sent = n()) %>%
  arrange(desc(n_sent))
print(emails_by_status_clean)

emails_by_status_clean <- message_with_status %>%
  filter(!is.na(status), status != "N/A") %>%
  group_by(status) %>%
  summarise(n_sent = n()) %>%
  arrange(desc(n_sent))
print(emails_by_status_clean)

ggplot(emails_by_status_clean, aes(x = reorder(status, n_sent), y = n_sent)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Emails Sent by Employee Role (Known Status Only)",
    x = "Employee Role (Status)",
    y = "Number of Emails Sent"
  ) +
  theme_minimal()


```
Observation: Employees and Vice Presidents were the most active communicators.




## Step 5: Monthly Email Volume

```{r step5, echo=FALSE, message=FALSE}
message <- message %>%
  mutate(date = as.POSIXct(date, tz = "UTC"))

message_clean <- message %>%
  filter(date >= as.POSIXct("1998-01-01", tz = "UTC") & date < as.POSIXct("2003-01-01", tz = "UTC"))

emails_over_time <- message_clean %>%
  mutate(month_str = format(date, "%Y-%m"),
         month = lubridate::ym(month_str)) %>%
  group_by(month) %>%
  summarise(n_emails = n(), .groups = "drop")

ggplot(emails_over_time, aes(x = month, y = n_emails)) +
  geom_line(color = "tomato", linewidth = 1) +
  labs(
    title = "Monthly Email Volume at Enron",
    subtitle = "Clear trends visible leading up to and during the 2001 Enron scandal",
    x = "Month",
    y = "Number of Emails Sent"
  ) +
  scale_x_date(
    date_labels = "%b %Y",
    date_breaks = "3 months"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
Insight: Email volume shows clear trends leading up to and during the 2001 Enron scandal.




## Step 6: Subject Line Word Frequency

```{r step6, echo=FALSE, message=FALSE}

subject_words <- message %>%
  select(subject) %>%
  filter(!is.na(subject)) %>%
  unnest_tokens(word, subject)

data("stop_words")  
subject_words_clean <- subject_words %>%
  anti_join(stop_words, by = "word")

top_subject_words <- subject_words_clean %>%
  count(word, sort = TRUE)

top_subject_words %>%
  slice_max(n, n = 15) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top Words in Enron Email Subjects",
    x = "Word",
    y = "Frequency"
  ) +
  theme_minimal()


```
Observation: Frequent terms like “fw,” “meeting,” “gas,” and “update” reflecting operational discussions, internal coordination, and energy-related business communications within Enron.




## Step 7: Interactive Shiny App

To complement the static analysis, I developed a **Shiny web application** that enables users to interactively explore the Enron email dataset.

## Features of the App:
- Filter emails by **sender** or **employee role**
- View **email volume trends** over time
- Explore **frequent words** in email subject lines
- Visualize insights using interactive plots and tables

To run the app locally:

shiny::runApp("app.R")



## Conclusion:

- Email volume and forwarding activity increased significantly during Enron’s collapse, showing intensified internal communication.

- Middle-level employees, particularly Vice Presidents and managers, were the main communicators, highlighting their key role in information flow.

- Most email subject lines focused on internal meetings and updates, indicating an emphasis on coordination and operations.

- The R programming language was effective for network analysis, time series evaluation, and text mining, providing deep insights into communication patterns.


## README:

Tools and libraries used:

dplyr, ggplot2, lubridate, tidytext, shiny
